---
title: "Build and deploy a stroke prediction model using R"
date: "`r Sys.Date()`"
output: html_document
author: "Tushar Kumar"
---

# About Data Analysis Report

This RMarkdown file contains the report of the data analysis done for the project on building and deploying a stroke prediction model in R. It contains analysis such as data exploration, summary statistics and building the prediction models. The final report was completed on `r date()`.

**Data Description:**

According to the World Health Organization (WHO) stroke is the 2nd leading cause of death globally, responsible for approximately 11% of total deaths.

This data set is used to predict whether a patient is likely to get stroke based on the input parameters like gender, age, various diseases, and smoking status. Each row in the data provides relevant information about the patient.

# Task One: Import data and data preprocessing

## Load data and install packages

```{r}
#load the Library
library(tidyverse) # Loads readr,  dplyr, tidyr,stringr, etc.
library(readxl)    # Needs to be loaded separately for Excel
library(dplyr)     #important to check unique string
library(janitor)   #help clean the column name into camel font
library(fastDummies)

raw_data <- read_csv("D:/Stroke_prediction/stroke-data.csv")
#glimpse shows data , its type and examples 
clean_data=raw_data
glimpse(clean_data)
#convert the column name into camel font . improve code readibility

clean_data <- clean_data %>%
  clean_names()

#summary help to find null, unique and statistical values
summary(clean_data)

```

## Describe and explore the data

```{r}
#Check the N/A value as it was not found in na values
sum(clean_data$bmi=="N/A")


clean_data %>%
  count(bmi,sort=TRUE)


# Assuming your dataframe is 'my_data' and the column is 'bmi'
clean_data <- clean_data %>%
  mutate(
    # na_if() turns the text "N/A" into a real R missing value (NA)
    # as.numeric() forces the column back to numbers so we can do math on it
    bmi = as.numeric(na_if(bmi, "N/A"))
  )


#replace this with mean as the demographic value like bmi have normal distribution
clean_data <- clean_data %>%
  mutate(
    bmi = replace_na(bmi,mean(bmi,na.rm=TRUE))
  )

hist(clean_data$bmi)

hist(clean_data$bmi, 
     main = "Distribution of BMI",  # Adds a title
     xlab = "BMI",                  # Labels the X-axis
     col = "lightblue")

#it is found that the gender , residence , smoking type , work type and marriage consist of string . 
#for data profiling it is important to convert them into integer incoding.
clean_data %>%
  count(gender,sort=TRUE) #shows unique string and frequency

clean_data %>%
  count(residence_type,sort=TRUE)

clean_data %>%
  count(work_type,sort=TRUE)

clean_data %>%
  count(ever_married,sort=TRUE)

clean_data %>%
  count(smoking_status,sort=TRUE)

#change the row whose gender is other as Male

clean_data<-clean_data %>%
  mutate(gender=case_when(gender=="Other"~"Male",TRUE~gender))

#According to research quiting smoking help reduce the stroke risk by 34% . it should be numbered less than unknown
#encoding string column as research shows quiting is better than not availability about it .

my_data=clean_data


my_data <- my_data %>%
  dummy_cols(select_columns = "work_type", 
             remove_first_dummy = FALSE) # This drops the first alphabetical category

my_data <- my_data %>%
  dummy_cols(select_columns = "residence_type", 
             remove_first_dummy = FALSE)

my_data <- my_data %>%
  dummy_cols(select_columns = "smoking_status", 
             remove_first_dummy = FALSE) 
my_data <- my_data %>%
  
  
  mutate(marriage_encoded = if_else(ever_married == "Yes", 1, 0))

my_data <- my_data %>%
  
  
  mutate(marriage_encoded = if_else(ever_married == "Yes", 1, 0))



# Check if gender matters when you also factor in smoking
my_data %>%
  group_by(gender, smoking_status) %>%
  summarise(
    total_people = n(),
    stroke_rate = (sum(stroke == 1, na.rm = TRUE) / n()) * 100
  )

#one hot encoding as both gender have specific category that have fluctuation in data
my_data <- my_data %>%
  dummy_cols(select_columns = "gender", 
             remove_first_dummy = TRUE) 

summary(my_data)

# 1. Convert the stroke column to a factor with meaningful labels
my_data <- my_data %>%
  mutate(stroke = factor(stroke, 
                         levels = c(0, 1), 
                         labels = c("No", "Yes")))





  





```

# Task Two: Build prediction models

```{r}
#after addressing the data profiling by converting null and encoding the string 
#Now we will procced to build the prediction models . By first spliting the data into 80-20 for train and split
library(tidymodels)
set.seed(123)

# strokes vs non-strokes in both sets.
data_split <- initial_split(my_data, prop = 0.8, strata = stroke)

train_data <- training(data_split)
test_data  <- testing(data_split)

# 1. Build the recipe using the training data
stroke_recipe <- recipe(stroke ~ ., data = train_data) %>%
  
  # Step 1: Explicitly remove the original raw text/string columns 
  step_rm(id,gender, work_type, ever_married, smoking_status,residence_type) %>%
  
  # Step 2: Handle any missing numeric values 
  step_impute_mean(all_numeric_predictors()) %>%
  
  # Step 3: Normalize all numeric and encoded columns so they are on the same scale
  step_normalize(all_numeric_predictors())



#To improve the accuracy of model. i will apply three model and keep the best model

#Linear regression model
log_reg_model <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

# 2. Create the workflow object
stroke_workflow <- workflow() %>%
  add_recipe(stroke_recipe) %>%  #  recipe with step_normalize()
  add_model(log_reg_model)       # The model just defined

stroke_fit <- stroke_workflow %>%
  fit(data = train_data)

#shows the coefficient for the regression model
stroke_fit


#random forest model
rf_model <- rand_forest(trees = 500) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

# 2. Create the workflow
rf_workflow <- workflow() %>%
  add_recipe(stroke_recipe) %>% # Reusing your existing recipe!
  add_model(rf_model)

# 3. Train the model
rf_fit <- rf_workflow %>%
  fit(data = train_data)

# Print a quick summary
rf_fit

# Neural Network Model
# hidden_units = number of nodes in the hidden layer
# penalty = regularization to prevent overfitting
nn_model <- mlp(hidden_units = 5, penalty = 0.01) %>%
  set_engine("nnet") %>%
  set_mode("classification")

# 2. Create the workflow
nn_workflow <- workflow() %>%
  add_recipe(stroke_recipe) %>%
  add_model(nn_model)

# 3. Train the model
# (This might take a few seconds depending on your machine)
nn_fit <- nn_workflow %>%
  fit(data = train_data)

# Print a quick summary
nn_fit

#Now is the best time to test the model predictibility and determine there accuracy and adapt to the best model

#let us start with logistical regression model
#renaming help to remember the predicted model
log_reg_preds <- predict(stroke_fit, new_data = test_data) %>%
  rename(pred_log_reg = .pred_class)

# 2. Predict using Random Forest
rf_preds <- predict(rf_fit, new_data = test_data) %>%
  rename(pred_rf = .pred_class)

# 3. Predict using Neural Network
nn_preds <- predict(nn_fit, new_data = test_data) %>%
  rename(pred_nn = .pred_class)

```

# Task Three: Evaluate and select prediction models

```{r}
#with all these model come the challenge to choose the one
#statistics offer the model to choose one 
#i am going to check the data on the basis of different #metrics including accuracy, recall, precision specificity

# Create the results dataframe
results_df <- test_data %>%
  select(actual_stroke = stroke) %>%
  bind_cols(log_reg_preds, rf_preds, nn_preds)

# Now check that this object exists:
head(results_df)
checking_metrics <- metric_set(accuracy, sens, spec, precision)

#let me start with logistic regression
checking_metrics(results_df, truth = actual_stroke, estimate = pred_log_reg)

#Random forest
checking_metrics(results_df, truth = actual_stroke, estimate = pred_rf)

#neural network
checking_metrics(results_df, truth = actual_stroke, estimate = pred_nn)

```

# Task Four: Deploy the prediction model

```{r}
# 1. SETUP
library(shiny)
library(tidymodels)
library(dplyr)

# Load the trained model (Ensure lg_model.rds is in the same folder as app.R)
lg_model <- readRDS("lg_model.rds")

# 2. USER INTERFACE (UI)
ui <- fluidPage(
  titlePanel("Stroke Risk Prediction Dashboard"),
  
  sidebarLayout(
    sidebarPanel(
      h4("Patient Demographics"),
      sliderInput("age", "Age:", min = 0, max = 100, value = 50),
      selectInput("gender", "Gender:", choices = c("Male", "Female")),
      selectInput("ever_married", "Ever Married:", choices = c("Yes", "No")),
      
      h4("Health Metrics"),
      numericInput("bmi", "BMI (Body Mass Index):", value = 28.5, min = 10, max = 80),
      numericInput("avg_glucose_level", "Average Glucose Level:", value = 105.0, min = 50, max = 300),
      selectInput("hypertension", "Hypertension:", choices = c("No" = 0, "Yes" = 1)),
      selectInput("heart_disease", "Heart Disease:", choices = c("No" = 0, "Yes" = 1)),
      
      h4("Lifestyle & Background"),
      selectInput("smoking_status", "Smoking Status:", 
                  choices = c("never smoked", "formerly smoked", "smokes", "Unknown")),
      selectInput("work_type", "Work Type:", 
                  choices = c("Private", "Self-employed", "Govt_job", "children", "Never_worked")),
      selectInput("residence_type", "Residence Type:", 
                  choices = c("Urban", "Rural")),
      
      br(),
      actionButton("predict_btn", "Calculate Risk", class = "btn-primary", width = "100%")
    ),
    
    mainPanel(
      h3("Prediction Results"),
      hr(),
      h4(textOutput("prediction_text")),
      h4(textOutput("probability_text")),
      br(),
      p("Note: This is an analytical tool based on machine learning, not medical advice.")
    )
  )
)

# 3. SERVER LOGIC
server <- function(input, output) {
  
  prediction_results <- eventReactive(input$predict_btn, {
    
    # Translate UI inputs into the EXACT dataframe the model was trained on
    # check.names = FALSE is critical here so R doesn't remove the spaces in the smoking status column names!
    new_patient <- data.frame(
      # Raw Numeric Inputs
      age = input$age,
      bmi = input$bmi,
      avg_glucose_level = input$avg_glucose_level,
      hypertension = as.numeric(input$hypertension),
      heart_disease = as.numeric(input$heart_disease),
      
      # Encoded Marriage & Gender
      marriage_encoded = ifelse(input$ever_married == "Yes", 1, 0),
      gender_Male = ifelse(input$gender == "Male", 1, 0),
      
      # Encoded Work Type
      work_type_children = ifelse(input$work_type == "children", 1, 0),
      work_type_Govt_job = ifelse(input$work_type == "Govt_job", 1, 0),
      work_type_Never_worked = ifelse(input$work_type == "Never_worked", 1, 0),
      work_type_Private = ifelse(input$work_type == "Private", 1, 0),
      `work_type_Self-employed` = ifelse(input$work_type == "Self-employed", 1, 0),
      
      # Encoded Residence Type
      residence_type_Rural = ifelse(input$residence_type == "Rural", 1, 0),
      residence_type_Urban = ifelse(input$residence_type == "Urban", 1, 0),
      
      # Encoded Smoking Status
      `smoking_status_formerly smoked` = ifelse(input$smoking_status == "formerly smoked", 1, 0),
      `smoking_status_never smoked` = ifelse(input$smoking_status == "never smoked", 1, 0),
      smoking_status_smokes = ifelse(input$smoking_status == "smokes", 1, 0),
      smoking_status_Unknown = ifelse(input$smoking_status == "Unknown", 1, 0),
      
      # --- CRITICAL FIX ---
      # Your recipe has step_rm(id, gender, work_type, ever_married, smoking_status, residence_type)
      # Therefore, we MUST include these original columns so the recipe has something to delete!
      id = 99999,
      gender = input$gender,
      ever_married = input$ever_married,
      work_type = input$work_type,
      residence_type = input$residence_type,
      smoking_status = input$smoking_status,
      
      check.names = FALSE 
    )
    
    # Get predictions
    pred_class <- predict(lg_model, new_data = new_patient)$.pred_class
    pred_prob <- predict(lg_model, new_data = new_patient, type = "prob")
    
    list(class = pred_class, prob = pred_prob$.pred_Yes)
  })
  
  # 4. OUTPUT TO UI
  output$prediction_text <- renderText({
    paste("Model Prediction:", prediction_results()$class)
  })
  
  output$probability_text <- renderText({
    prob_pct <- round(prediction_results()$prob * 100, 2)
    paste("Probability of Stroke:", prob_pct, "%")
  })
}

# 5. RUN THE APP
shinyApp(ui = ui, server = server)

```

# Task Five: Findings and Conclusions

Core Analytical InsightsStroke is a highly complex medical event driven by multifaceted risk factors. Developing a robust predictive model required rigorous exploratory data analysis (EDA) and visualization to uncover underlying correlations. The analysis clearly identified that clinical metrics—specifically hypertension, existing heart disease, and average glucose levels—are the strongest primary predictors of stroke risk. Furthermore, these physiological factors are heavily augmented by lifestyle elements, including work type and smoking status.Data Preprocessing & Feature EngineeringTo ensure the models could effectively interpret this complexity, the raw data required a highly structured preprocessing pipeline:Data Cleaning: Column names were standardized for programmatic consistency. The bmi feature contained missing values recorded as "N/A" character strings, which were converted to true nulls and imputed using the statistical mean to preserve the normal distribution. Additionally, extreme minority classes (such as a single "Other" entry in the gender column) were addressed to prevent algorithm errors.Encoding: Categorical variables (residence type, marital status, smoking history, and work type) were converted using dummy encoding. While aggregate demographic features like gender showed minimal variance on the surface, interaction analysis revealed deeper subgroup significance, making categorical encoding a critical step.Target Variable: The dependent variable (stroke) was transformed into a binary Factor (Yes/No) to properly set up the classification models.Model Evaluation & PerformanceWith a fully normalized and encoded dataset, three distinct predictive architectures were trained and evaluated: Logistic Regression, a Random Forest, and a Neural Network (Multi-Layer Perceptron).Upon evaluating the testing data using standard clinical metrics, Logistic Regression achieved the best overall performance—scoring the highest in both Accuracy and Recall (Sensitivity).Random Forest performed adequately as a runner-up, while the Neural Network showed that simpler models often perform better on structured tabular data.Future Scope & DeploymentWhile the current Logistic Regression model demonstrates strong predictive capabilities, further hyperparameter tuning and alternative imputation strategies for missing fields could theoretically increase the model's sensitivity.As a final step in the project lifecycle, this predictive algorithm is being deployed from the local R environment into a live, interactive web dashboard. This transitions the project from a static analysis into a real-world, customer-facing application.
